Ecological Limits to Forecasting: Estimating Uncertainty in Critical Transitions with Deep Learning

### 2022-05-13 manuscript/manuscript.Rmd

We provide fully reproducible coded examples in R and python for performing, scoring, and visualizing each of the forecasts considered here.
After considerable time considering alternative frameworks, we have emphasized those which best met our requirements for performance, ease-of-use, flexibility, and support for the latest probabilistic machine learning models for forecasting. 
Most of our forecasts use the `darts` framework, a sophisticated and well documented python library with support for a wide range of methods.
Our model-based MCMC forecasts use the `greta` framework, an R library that uses python-based `tensorflow` probability to achieve better performance.
While python-based frameworks currently have the edge in performance and access modern ML algorithms, they lag behind in attention to statistical issues such as the computation of strictly proper skill scores (discussed below.) 
Thus our examples of scoring and visualization will rely on a collection of R packages, in particular, `scoringRules` for the efficient calculation of Continuous Ranked Probability Score (CRPS) and logarithmic probability scores for forecast ensembles [@gneiting_strictly_2007].
We expect greater convergence between methods available in R and python in the future, as already illustrated in the example of `greta`.
Complete code for all examples presented here can be found at https://github.com/boettiger-lab/mee_tipping_point_forecasting.

@@

Following @dakos_methods_2012, we further allow the carrying capacity of the host, $K$ to slowly increase at linear rate, which drives a Hopf bifurcation as $K$ becomes sufficiently large. 
In a Hopf bifurcation, a stable node starts an oscillatory pattern which continues to grow in amplitude as the bifurcation parameter continues to increase.
This example illustrates one of the many kinds of challenges which nonlinear phenomena pose to forecasting: the "historical" data prior to the bifurcation never exhibit the cyclical dynamics of growing amplitude that will emerge after the bifurcation occurs.
If we had used a purely deterministic model, the dynamics would be constrained to a single stable point, corresponding to a slowly changing steady-state population size of host and parasitoid populations.
However, stochasticity in this case acts as a source of some additional information about the dynamics, as the noise excites quasi-cycles which are visible in the irregular oscillations that appear significantly prior to the emergence of true limit cycles which follow the bifurcation [@boettiger_noise_2018].

@@

If the environment slowly alters one of the parameters (say, the encounter efficiency, `h_t`, in our formulation), one of the stable nodes moves closer and closer to the unstable saddle point, leading to a bifurcation that destroys the stable state, leaving the system to suddenly transition to the alternative stable state.
Saddle-node bifurcations (also known as fold bifurcations) also create a phenomenon known as hysteresis, where it is not sufficient to restore the environment to the previous parameter values to recover the previous state. 
Unlike the Hopf bifurcation which exhibits a continuous transition from a stable node to a small limit cycle that then grows, the saddle-node transition is a discontinuous or so-called 'catastrophic' bifurcation.
Due both to this sudden, catastrophic nature of the transition and the difficulty in reversing the shift after it has occurred, saddle-node bifurcations have been the subject of intense study.
Tipping point dynamics have long been identified as an important but difficult challenge for forecasting [e.g. @scheffer_catastrophic_2001; @Folke2004]. 
Much effort in the ecological literature so far has focused on identifying any 'early warning signs' that a catastrophic bifurcation might occur at all [@scheffer_early-warning_2009; @scheffer_anticipating_2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs.
Tipping points resulting from Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dai2013; @Dai2015] to whole-ecosystem experiments [@carpenter_early_2011], and postulated as a model for global change [@Barnosky2012].

@@

In this context, predicting the probability of a transition in the future based solely on observations prior to a transition occurring is essentially impossible without additional information constraining the model estimate, as such data is equally consistent with infinitely many models or parameter choices which share the same local linearization about the stable point.  
Unlike the saddle-node bifurcation, there is no slowly warping potential basin which can be detected to inform estimates. 
Thus, in this scenario, rather than considering the problem of predicting the future evolution of a single time series based only on it's historical values, we consider an alternative framing of the task:
we imagine our forecaster has access to historical data from one or more comparable systems which includes a previous stochastic transition event.
Based on this data, our forecaster seeks to identify the distribution of expected transition times for analagous systems starting from the same initial condition. 
This parallels actual practice in which researchers would draw on previous examples of stochastic transitions in a system - lake-ecosystem shifts, disease emergence, changing fire regimes, [@Scheffer2001; @Folke2004].
(Note that such stochastic transitions between alternative stable states can also create oscillatory-like dynamics when stochasticity is sufficiently high enough to drive repeated transitions from one attractor to the other and back again.  In such cases, it might be reasonable to estimate a strictly forward-looking forecast of a single system, predicting the distribution of these transitions.)

@@

<!-- Key distinction here from MCMC above is not on how model parameters are estimated, but rather on approaches that aren't making strong mechanistic assumptions about process, but are merely phenomonological type models
-->

@@

@@

>>>>

We provide fully reproducible coded examples in R and python for performing, scoring, and visualizing each of the forecasts considered here.
After considerable time considering alternative frameworks, we have emphasized those which best met our requirements for performance, ease-of-use, flexibility, and support for the latest probabilistic machine learning models for forecasting. 
Most of our forecasts use the `darts` framework, a sophisticated and well documented python library with support for a wide range of methods.
Our model-based MCMC forecasts use the `greta` framework, an R library that uses python-based `tensorflow` probability to achieve better performance.
While python-based frameworks currently have the edge in performance and access modern ML algorithms, they lag behind in attention to statistical issues such as the computation of strictly proper skill scores (discussed below.)

Thus our examples of scoring and visualization will rely on a collection of R packages, in particular, `scoringRules` for the efficient calculation of Continuous Ranked Probability Score (CRPS) and logarithmic probability (Logs) scores for forecast ensembles [@gneiting_strictly_2007].
Following popular conventions, we express both skill scores in error-orientation, that is, larger values indicate worse skill (higher degree of error).

We expect greater convergence between methods available in R and python in the future, as already illustrated in the example of `greta`.
Complete code for all examples presented here can be found at https://github.com/boettiger-lab/mee_tipping_point_forecasting.

@@

Following @dakos_methods_2012, we further allow the carrying capacity of the host, $K$ to slowly increase at linear rate, which drives a Hopf bifurcation as $K$ becomes sufficiently large. 
In a Hopf bifurcation, a stable node starts an oscillatory pattern which continues to grow in amplitude as the bifurcation parameter continues to increase.
This example illustrates one of the many kinds of challenges which nonlinear phenomena pose to forecasting: the "historical" data prior to the bifurcation never exhibit the cyclical dynamics of growing amplitude that will emerge after the bifurcation occurs.
If we had used a purely deterministic model, the dynamics would be constrained to a single stable point, corresponding to a slowly changing steady-state population size of host and parasitoid populations.
However, stochasticity in this case acts as a source of some additional information about the dynamics, as the noise excites quasi-cycles which are visible in the irregular oscillations that appear significantly prior to the emergence of true limit cycles which follow the bifurcation [@boettiger_noise_2018].
Examples use the following parameters: $H_0 = 9$, $P_0 = 1$, $r = 0.75$, $c = 0.1$, $K_0 = 14$, $\delta = 0.08$, $\sigma_H = 0.02$, $\sigma_P = 0.02$.

@@

If the environment slowly alters one of the parameters (say, the encounter efficiency, `h_t`, in our formulation), one of the stable nodes moves closer and closer to the unstable saddle point, leading to a bifurcation that destroys the stable state, leaving the system to suddenly transition to the alternative stable state.
Saddle-node bifurcations (also known as fold bifurcations) also create a phenomenon known as hysteresis, where it is not sufficient to restore the environment to the previous parameter values to recover the previous state. 
Unlike the Hopf bifurcation which exhibits a continuous transition from a stable node to a small limit cycle that then grows, the saddle-node transition is a discontinuous or so-called 'catastrophic' bifurcation.
Due both to this sudden, catastrophic nature of the transition and the difficulty in reversing the shift after it has occurred, saddle-node bifurcations have been the subject of intense study.

Tipping point dynamics have long been identified as an important but difficult challenge for forecasting [e.g. @Scheffer2001; @Folke2004]. 
Much effort in the ecological literature so far has focused on identifying any 'early warning signs' that a catastrophic bifurcation might occur at all [@Scheffer2009; @Scheffer2012] rather than more ambitious attempts to provide quantitative probabilistic forecasts of the likely distribution of waiting times before such a transition occurs.
Tipping points resulting from Saddle-node bifurcations have been demonstrated in examples ranging from laboratory microcosms [@Dai2012; @Dai2015] to whole-ecosystem experiments [@carpenter_early_2011], and postulated as a model for global change [@Barnosky2012]. 
Examples use the following parameters: $r = 1$, $K = 1$, $s = 0.1$, $h_0 = 0.15$, $\alpha = 0.000375$, $\sigma = 0.02$, $N_0 = 0.75$.

@@

In this context, predicting the probability of a transition in the future based solely on observations prior to a transition occurring is essentially impossible without additional information constraining the model estimate, as such data is equally consistent with infinitely many models or parameter choices which share the same local linearization about the stable point.  
Unlike the saddle-node bifurcation, there is no slowly warping potential basin which can be detected to inform estimates. 
Thus, in this scenario, rather than considering the problem of predicting the future evolution of a single time series based only on it's historical values, we consider an alternative framing of the task:
we imagine our forecaster has access to historical data from one or more comparable systems which includes a previous stochastic transition event.
Based on this data, our forecaster seeks to identify the distribution of expected transition times for analagous systems starting from the same initial condition. 
This parallels actual practice in which researchers would draw on previous examples of stochastic transitions in a system - lake-ecosystem shifts, disease emergence, changing fire regimes, [@Scheffer2001; @Folke2004].
(Note that such stochastic transitions between alternative stable states can also create oscillatory-like dynamics when stochasticity is sufficiently high enough to drive repeated transitions from one attractor to the other and back again.  In such cases, it might be reasonable to estimate a strictly forward-looking forecast of a single system, predicting the distribution of these transitions.)
Model definition is the same as May's model for the saddle node with fixed parameter $h$, values: $r = 1$, $K = 1$, $s = 0.1$, $h_0 = h = 0.26$, $\alpha = 0$, $\sigma = 0.02$, $N_0 = 0.55$.

@@

<!-- Key distinction here from MCMC above is not on how model parameters are estimated, but rather on approaches that aren't making strong mechanistic assumptions about process, but are merely phenomenological type models
-->

@@

Our analysis has focused on the task of making a single forecast prior to the occurrence of a critical transition.
Forecasting is ideally a more iterative process of data assimilation, where forecasts are updated with respect to additional observations, rather than projecting 100s of time steps into the future [@Dietz2018]. 
Updating a forecast after a critical transition has already occurred may be of little use in the context of hysteresis, such as under the saddle node or stochastic transition -- recognizing the alternative stable state only after the system is stuck in that basin will often be considered 'too late'.
Assimilation may be more applicable to the Hopf bifurcation, where additional observations of slowly growing oscillations may lead to more accurate forecasts. 
Such models may even accurately predict the homoclinic bifurcation that occurs when the limit cycle grows too large, eventually hitting a saddle point of zero population size for the host species.
We leave these cases to future exploration.

@@

- Forecasts with 'assimilation' i.e. ability to re-estimate after seeing some of the data, rather than looking at a single forecast with a long horizon.
